The Concurrent Calculi Formalisation Benchmark

### 2024-02-19 short-paper/short-paper.tex
Small edits for brevity and some spelling fixes.

The POPLMark challenge~\cite{POPLMark}
played a pivotal role in advancing the field of proof assistants,
libraries, and best practices for programming language
mechanization. By providing a shared framework for systematically
evaluating mechanization techniques, it catalyzed a significant
shift within the programming language research community. As a
result, scholarly publications increasingly incorporate
mechanized proofs.
%
% The POPLMark challenge~\cite{POPLMark} accelerated the development of proof assistants, libraries, and best practices for programming language mechanisation by establishing a common ground for systematically evaluating mechanisation techniques.
% In doing so, it spearheaded a shift in the programming language research community towards publications that include mechanised proofs.
% POPLMark Reloaded~\cite{POPLMarkReloaded} proposed a similar program for proofs involving logical relations.
% These challenges, and the developments they spurred, were limited in scope, but their authors noted the need to tackle topics such as coinduction and linearity in the future.

@@

In this spirit, we introduce a novel collection of benchmark problems
specifically crafted to tackle common challenges encountered during
the mechanization of formal models for concurrent and distributed
programming languages. Our focus extends particularly to process
calculi.
%
% In this spirit, we propose a new collection of benchmark problems
% specifically designed to address typical issues that arise when
% mechanising formal models of concurrent and distributed programming
% languages in general, and process calculi in particular. 
Concurrent calculi are notably subtle: For instance, it took nine
years before the incorrect proof of type safety in the seminal paper
by Honda et al.~.~\cite{Honda1998} was rectified in Yoshida \&
Vasconcelos’s subsequent work\cite{Yoshida2007}.
% for instance it took nine years before the incorrect proof of type
% safety in the seminal paper by Honda et al was fixed
% in~\cite{Yoshida2007}.
Similarly, many
key results in peer-reviewed papers on session types have turned out
to be false later on~\cite{Gay2020,10.1145/3290343}, illustrating the
value of machine-checkable proofs.

While results about concurrent
formalisms have already been mechanised (as we will discuss further
below), our experience is that choosing appropriate mechanisation
techniques and tools remains a significant challenge and that the
trade-offs are not well understood.  This often leads researchers
toward a trial-and-error approach, resulting in sub-optimal solutions,
wasted mechanisation effort, and techniques and results that are hard
to reuse.  For example, the authors in~\cite{Cruz-Filipe2021b} note that the high-level parts of
mechanised proofs closely resemble the informal versions, and that the
main challenge lies in getting the foundations right.

Our benchmark problems (explained in detail in \cref{app:challenges}) center around three critical aspects that frequently pose challenges in mechanizing concurrency theory:
% focus on three key aspects which often cause difficulties when mechanising concurrency theory:
\emph{linearity}, \emph{scope extrusion}, and \emph{coinductive
  reasoning}.  Mechanisations must often address several of these
aspects at the same time; however, we see the combination of
techniques as a next step beyond the scope of this challenge, as
discussed further in \Cref{sec:going-beyond}.  Our benchmark
challenges are based on process calculi, as these provide a simple but
realistic showcase of common issues encountered during mechanization endeavors
% exhibit of typical issues encountered in mechanisation work.
While the three highlighted aspects are not the sole sources of difficulties, they
consistently manifest in most mechanizations of concurrency theory.
% The three aspects are certainly not the only causes of difficulties,
% but as we will discuss presently, they emerge in most mechanisations
% of concurrency theory.

@@

\paragraph{Coinduction.} \todo{explain more what the challenge does.}
\lore{I have tried this:
  the first paragraph motivates (``why is this difficult?''), while
  the second talks about the challenge.}
Coinduction is fundamental to establish properties of processes that
may never terminate, which is a typical situation.
Common uses for coinduction are to define strong and weak versions of
bisimulation, barbed notions of bisimulation, and prove trace equivalence. %,
%etc.
On the other hand, coinduction in proof assistants
is notoriously difficult and its representation
is non-uniform across development.
E.g., \cite{Castro-Perez2021} notes that the most challenging
part of their mechanisation of multiparty session types
was the convoluted definition of infinite behaviours;
\cite{Pohjola2022} points out that the main difficulty
of the proofs concerning bisimulation lies
in dealing with existential quantifiers; and
\cite{Bengtson2016} wonders whether parametrised
coinduction as developed in Coq \cite{Hur2013} would
have helped their Isabelle mechanisation.

@@

Still, these (and other) applications need the basic
techniques that our challenge problems emphasize:
Aligned with design goal~\ref{item:goal-reusability}, our problems are rooted in the fundamental meta-theory of process calculi, without necessitating the creation of novel theory. These problems center around well-established theorems and results, showcasing intriguing proof techniques that can be leveraged in future endeavors.
% as per design goal
% \ref{item:goal-reusability}, our problems are drawn from the basic
% meta-theory of process calculi (without requiring the development of
% new theory), and focus on well-known theorems and results that involve
% interesting proof techniques that may be reused in further work. 
To
achieve design goal \ref{item:goal-comperison-accessibility}, we have
formulated our challenge problems to explore the three aspects
(linearity, scope extrusion, coinduction) independently, so that they
may be solved individually and in any order; each problem is %should be
reasonably small and easily understandable with basic knowledge of
textbook concurrency theory, process calculi, and type theory.  The
process calculus used in each challenge focuses on the features that
we want to emphasize, and omits all constructs (such as choices) that
would complicate the mechanisation without bringing additional
tangible benefits and insights.  The minimality and uniformity of the
calculi also allows us to target design goal
\ref{item:goal-tutorials}.

@@

\alb{This should not be a subsubsection at the same level of coinduction etc}
\subsubsection{Evaluation criteria.}
The idea behind our benchmark is to obtain evidence towards answering
questions \cref{item:rq1,item:rq2,item:rq3}. We are therefore
interested not only in the solutions, but also in the experience of
solving the challenges with the chosen approach.  Solutions to our
challenges should be compared on three axis:
\begin{enumerate}
\item Mechanisation overhead: the amount of manually written infrastructure and setup needed to express the definitions in the mechanisation;
\item Adequacy of the formal statements in the mechanisation: whether the proven theorems are easily recognisable as the theorems from the challenge; and
\item Cost of entry for the tools and techniques employed: the difficulty of learning to use the techniques.
\end{enumerate}
Solutions to our challenges do not need to strictly follow the definitions and lemmata set out in the problem text, but solutions which deviate from the original challenges must present more elaborate argumentation for their adequacy.

>>>>

The POPLMark challenge~\cite{POPLMark}
played a pivotal role in advancing the field of proof assistants,
libraries, and best practices for programming language
mechanisation. By providing a shared framework for systematically
evaluating mechanisation techniques, it catalyzed a significant
shift towards publications that include mechanised proofs within the programming language research community.

@@

In this spirit, we introduce a new collection of benchmark problems
specifically crafted to tackle common challenges encountered during
the mechanisation of formal models for concurrent and distributed
programming languages. Our focus extends particularly to process
calculi.
Concurrent calculi are notably subtle: For instance, it took nine
years before the incorrect proof of type safety in the seminal paper
by Honda et al.~\cite{Honda1998} was rectified in Yoshida \&
Vasconcelos’s subsequent work\cite{Yoshida2007}.
Similarly, many
key results in papers on session types have turned out
to be false later on~\cite{Gay2020,10.1145/3290343}, illustrating the
value of machine-checkable proofs.

While results about concurrent
formalisms have already been mechanised (as we will discuss further
below), our experience is that choosing appropriate mechanisation
techniques and tools remains a significant challenge and that the
trade-offs are not well understood.  This often leads researchers
toward a trial-and-error approach, resulting in sub-optimal solutions,
wasted mechanisation effort, and techniques and results that are hard
to reuse. For example, the authors in~\cite{Cruz-Filipe2021b} note that the high-level parts of
mechanised proofs closely resemble the informal ones, and that the
main challenge lies in getting the foundations right.

Our benchmark problems (detailed in \cref{app:challenges}) center around three key aspects that frequently pose challenges when mechanising concurrency theory: \emph{linearity}, \emph{scope extrusion}, and \emph{coinductive reasoning}.
Mechanisations must often address several of these
aspects at the same time; however, we see the combination of
techniques as a next step beyond the scope of this challenge, as
discussed further in \Cref{sec:going-beyond}.
Our benchmark challenges are based on process calculi, as these provide a simple but
realistic showcase of common issues encountered during mechanisation efforts.
While the three highlighted aspects are not the sole sources of difficulties, they emerge in most mechanisations of concurrency theory.

@@

\paragraph{Coinduction.} \todo{explain more what the challenge does.}
\lore{I have tried this:
  the first paragraph motivates (``why is this difficult?''), while
  the second talks about the challenge.}
Coinduction is fundamental to establish properties of processes that
may never terminate, which is a typical situation.
Common uses for coinduction are to define strong and weak versions of
bisimulation, barbed notions of bisimulation, and prove trace equivalence.
On the other hand, coinduction in proof assistants
is notoriously difficult and its representation
is non-uniform across developments.
E.g., \cite{Castro-Perez2021} notes that the most challenging
part of their mechanisation of multiparty session types
was the convoluted definition of infinite behaviours;
\cite{Pohjola2022} points out that the main difficulty
of the proofs concerning bisimulation lies
in dealing with existential quantifiers; and
\cite{Bengtson2016} wonders whether parametrised
coinduction as developed in Coq \cite{Hur2013} would
have helped their Isabelle mechanisation.

@@

Still, these (and other) applications need the basic
techniques that our challenge problems emphasize:
Aligned with design goal~\ref{item:goal-reusability}, our problems are rooted in the fundamental meta-theory of process calculi, without necessitating the creation of novel theory. These problems center around well-established theorems and results, showcasing interesting proof techniques that can be leveraged in future endeavors.
% as per design goal
% \ref{item:goal-reusability}, our problems are drawn from the basic
% meta-theory of process calculi (without requiring the development of
% new theory), and focus on well-known theorems and results that involve
% interesting proof techniques that may be reused in further work. 
To
achieve design goal \ref{item:goal-comperison-accessibility}, we have
formulated our challenge problems to explore the three aspects
(linearity, scope extrusion, coinduction) independently, so that they
may be solved individually and in any order; each problem is %should be
reasonably small and easily understandable with basic knowledge of
textbook concurrency theory, process calculi, and type theory.  The
process calculus used in each challenge focuses on the features that
we want to emphasize, and omits all constructs (such as choices) that
would complicate the mechanisation without bringing additional
tangible benefits and insights.  The minimality and uniformity of the
calculi also allows us to target design goal
\ref{item:goal-tutorials}.

@@

\alb{This should not be a subsubsection at the same level of coinduction etc}
\subsubsection{Evaluation criteria.}
The idea behind our benchmark is to obtain evidence towards answering
questions \cref{item:rq1,item:rq2,item:rq3}. We are therefore
interested not only in the solutions, but also in the experience of
solving the challenges with the chosen approach.  Solutions to our
challenges should be compared on three axes:
\begin{enumerate}
\item Mechanisation overhead: the amount of manually written infrastructure and setup needed to express the definitions in the mechanisation;
\item Adequacy of the formal statements in the mechanisation: whether the proven theorems are easily recognisable as the theorems from the challenge; and
\item Cost of entry for the tools and techniques employed: the difficulty of learning to use the techniques.
\end{enumerate}
Solutions to our challenges do not need to strictly follow the definitions and lemmata set out in the problem text, but solutions which deviate from the original challenges must present more elaborate argumentation for their adequacy.
